---
title: "Homework Assignment 4"
subtitle: "Prioritizing potential aquaculture" 
date: last-modified
execute: 
  eval: true
  message: false
  warning: false
format:
  html:
    toc: true
editor_options: 
---

## Background

Aquaculture is a quickly growing field as a means to supply more sustainable protein options without the land use. This assignment aims to determine which Exclusive Economic Zones (EEZ) on the West Coast are best suited as marine aquaculture sites for oysters and sugar kelp (Saccharina latissima). Sites are choosen based on a range of suitable sea surface temperatures and depth values for the particular species.

## Set Up

### Libraries

```{r}
#| code-fold: true
#| code-summary: "Show the code"
library(tidyverse)
library(dplyr)
library(here)
library(sf)
library(terra)
library(stars)
library(tmap)
library(raster)
library(spData)
library(kableExtra)
library(testthat)
```

### Data

Sea surface temperature data from 2008 to 2012 generated from NOAA's 5km Daily Satellite Sea Surface Temperature Anomaly v3.1 \[1\]

Bathymetry data comes from General Bathymetric Charts of the Oceans (GEBCO). \[2\]

The designation of Exclusive Economic Zones off the West Coast comes from Marineregions.org, which outlines the boundaries of marine biogeographic areas. \[3\] The names of the regions are: Washington, Oregon, Northern California, Central California, and Southern California.

#### SST and depth requirements

Oysters:

-   optimal sea surface temperature: 11-30째C
-   depth: 0-70 meters below sea level

Sugar Kelp \[4\]:

-   optimal sea surface temperature: 10-15째C
-   depth: 0-30 meters below sea level

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# ---- Raster stack of sea surface temperatures from 2008 to 2012
# Set directory 
sst_path <- here::here('data')

# Create list of all SST .tif files 
sst_files <- list.files(sst_path, 
                        pattern = glob2rx("*average_annual_sst_*.tif$"),
                        full.names = TRUE)
# Create a raster stack 
sst <- rast(sst_files) 

# ---- Bathymetry data 
depth <- terra::rast(here::here('data', 'depth.tif'))

# ---- EEZ 
eez <- read_sf(here::here('data', 'wc_regions_clean.shp'))

# ---- West coast states  
# load US states
us_states <- spData::us_states

# filter to ca, or, wa
west_coast_states <- us_states %>% 
  filter(NAME %in% c("California", "Oregon", "Washington"))
```

## Process Data

In order to find which area is suitable for oyster production in terms of sea surface temperature and depth, the data needs to be processed. The following code matches the coordinate reference systems (CRS) of all the spatial data, finds the mean sea surface over the time period, and converts the temperature from Kelvin to Celsius. Then, the resolution and extents of the raster data are alligned so that the depth and temperature data can be combined. 

#### CRS and Temp

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# ---- Coordinate Reference Systems
# Reproject CRS to all match sst
depth <- terra::project(depth, crs(sst))
eez <- st_transform(eez, crs(sst))
west_coast_states <- st_transform(west_coast_states, crs(sst))

#---- Finding mean sea surface temperatures from 2008-2012
# Mean of raster stack 
sst_mean <- mean(sst)

# ---- Convert SST from Kelvin to Celsius
# Convert to Celsius by subtracting 273.15 
sst_c <- sst_mean - 273.15

# Rename layer name 
names(sst_c) <- "temp_c"
```


#### Extent and Resolution

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| output: false
#| quiet: true
# ---- Crop depth raster to match the extent of SST raster

# View it before crop 
map1 <- tm_shape(depth) + 
  tm_raster() + 
  tm_layout(title = "Pre-crop")

# Crop depth extent to that of sst
depth_range <- terra::crop(depth, sst_c)

# View what happened
map2 <- tm_shape(depth_range) + 
  tm_raster() + 
  tm_layout(title = "Post-crop")

tmap_arrange(map1, map2, nrow = 1)

# ---- Resolution 
# Check if resolution of depth and sst match 
res(depth_range) == res(sst_c)

# Change resolution of depth to match sst using nearest neighbor approach
depth_rs <- terra::resample(depth_range, y = sst_c, method = "near")
```

Now let's confirm the dimensions of data match before merging them

```{r}
# Test to check rasters are compatible for stacking
test_that("Resolution, extent, and CRS are the same", {
  # Check resolution
  expect_equal(res(depth_rs), res(sst_c), label = "Resolution does not match")
  # Check extent
  expect_equal(ext(depth_rs), ext(sst_c), label = "Extent does not match")
  # Check CRS
  expect_equal(st_crs(depth_rs), st_crs(sst_c), label = "CRS does not match")
})
```

## Find Suitable locations for oyster conditions

In order to find cells within the raster that are suitable for oyster conditions, I will reclassify the raster as the following:

Set the following values to `1`

-   Depth: 0-70 meters
-   Sea surface temperature: 11-30째C

Any other value set to `0`

```{r}
#| code-fold: true
#| code-summary: "Show the code"
# ---- Reclassification 
# Create reclasification matrix for depth 
rcl_depth <- matrix(c(-Inf, -70, 0,  # values -Inf to -70 = 0
                      -70, 0, 1,     # values -70 to 0 = 1
                      0, Inf, 0),    # values 0 to Inf = 0
                    ncol = 3, byrow = TRUE)

# Create reclasification matrix for temp 
rcl_sst <- matrix(c(-Inf,11, 0,# values from -Inf to 11 = 0
                      11, 30, 1, # values 11 to 30 = 1
                      30, Inf, 0), # values 30 to Inf = 0
                    ncol = 3, byrow = TRUE)

# Reclassify the raster for oyster specifications 
depth_oyster <- terra::classify(depth_rs, rcl = rcl_depth)
sst_oyster <- terra::classify(sst_c, rcl = rcl_sst)

# ---- Raster stack 
# Create raster stack 
oyster_stack <- c(depth_oyster, sst_oyster)

# Apply multiplication 
suitable_oyster <- terra::lapp(oyster_stack, fun = function(depth, sst) depth * sst)

# Turn 0 values into NA 
suitable_oyster[suitable_oyster == 0] <- NA
```

## Determine the most suitable EEZ

With the suitable areas for oyster aquaculture identified, the next step is to determine how much of these areas fall within each Exclusive Economic Zone (EEZ) along the west coast. Building on the previous analyses, I will create a map and a table to visualize and summarize these findings.


```{r}
#| code-fold: true
#| code-summary: "Show the code"

# ---- Calculate % of suitable area per zone 

# Rasterize the eez dataframe using the 'rgn' column
eez_rast <- terra::rasterize(eez, suitable_oyster, field = "rgn")

# Create dataframe that displays suitable area of each zone
suitable_area <- terra::expanse(suitable_oyster, unit = "km", zones = eez_rast)

# Rename columns
suitable_area <- suitable_area %>% 
  rename("suitable_area_km2" = "area",
         "rgn" = "zone")

# Calculate percent of each zone that contains suitable area for oyster aquaculture 
eez_oyster <- eez %>% 
  left_join(suitable_area, by= "rgn") %>% 
  mutate(zone_pct = suitable_area_km2 / area_km2 * 100)

# ---- Base map 

# Create bounding box of extents  
bbox_depth <- as.matrix(terra::ext(depth_range))

basemap <- tm_shape(depth_range) + # depth raster to show borders 
  tm_raster(palette = "-GnBu", midpoint = NA, alpha = 0.4, legend.show = FALSE) +
tm_shape(west_coast_states, bbox = bbox_depth) + # state outlines 
  tm_borders(alpha = 0.5) + 
  
  # labels for eez zones 
  tm_credits("Washington", position = c(0.26, 0.82), size = 0.6) +
  tm_credits("Oregon", position = c(0.25, 0.68), size = 0.6) +
  tm_credits("Northern\nCalifornia", position = c(0.25, 0.49), size = 0.6) +
  tm_credits("Central\nCalifornia", position = c(0.38, 0.3), size = 0.6) +
  tm_credits("Southern\nCalifornia", position = c(0.58, 0.11), size = 0.6) +
  
  tm_scale_bar(position = c(0.06, 0.04)) + # scale bar and compass 
  tm_compass(position = c(0.03, 0.1), size = 1.5) 


# ---- Map for oyster aquaculture 
basemap + tm_shape(eez_oyster) +  # eez zones 
  tm_polygons(col = "zone_pct",
              palette = "Oranges", 
              alpha = 0.5,
              title = "% of suitable area \nper EEZ (km2)") + 
  tm_shape(suitable_oyster) + # suitable oyster area
  tm_raster(palette = "firebrick", legend.show = FALSE) +
  tm_add_legend(type = "fill", 
                title = "Total suitable area (km^2)", 
                col = "firebrick", 
                labels = " ") +
  
  tm_layout(legend.outside = TRUE,
            fontfamily = "serif",
            main.title = "Oyster Aquaculture Suitable Sites")

# ---- Table 
# Filter data for table 
eez_table <- eez_oyster %>% 
  st_drop_geometry() %>% 
  dplyr::select(rgn, area_km2, suitable_area_km2, zone_pct) 

kable(eez_table,
      col.names = c("EEZ Region",
                    "EEZ total area (km^2)",
                    "Suitable aquaculture area (km^2)",
                    "% of suitable area per zone"),
      caption = "Percentage of suitable oyster aquaculture sites per EEZ",
      digits = 2) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center", html_font='times new roman')
  
```

## Function to input other species 

The following code takes the analysis above and turns it into a function that allows for other species's depth and temperature specifications to find suitable aquaculture sites. Additionally, the function produces a map and a table to better visualize the results. 

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# ----- Aquaculture function 
aqua_fun <- function(depth_low, depth_high, temp_low, temp_high, species_name){
  
  # reclassification matrices
  rcl_depth <- matrix(c(-Inf, depth_low, 0,  
                      depth_low, depth_high, 1, # suitable range
                      depth_low, Inf, 0),    
                    ncol = 3, byrow = TRUE)
  rcl_sst <- matrix(c(-Inf,temp_low, 0,
                      temp_low, temp_high, 1, # suitable range
                      temp_high, Inf, 0),
                    ncol = 3, byrow = TRUE)
  # reclassify depth and temp rasters 
  aqua_depth <- terra::classify(depth_rs, rcl = rcl_depth)
  aqua_sst <- terra::classify(sst_c, rcl = rcl_sst)
  
  # raster stack 
  aqua_stack <- c(aqua_depth, aqua_sst)
  
  # suitable locations  
  suitable_aqua <- terra::lapp(aqua_stack, fun = function(depth, sst) depth * sst)

  # Turn 0 values into NA 
  suitable_aqua[suitable_aqua == 0] <- NA
  
  # Rasterize the eez dataframe using the 'rgn' column
  eez_rast <- terra::rasterize(eez, suitable_aqua, field = "rgn")

  # Create dataframe that displays suitable area of each zone
  suitable_area <- terra::expanse(suitable_aqua, unit = "km", zones = eez_rast)

  # Rename columns
  suitable_area <- suitable_area %>% 
    rename("suitable_area_km2" = "area", "rgn" = "zone")

  # Calculate percent of each zone that contains suitable area for oyster aquaculture 
  eez_aqua <- eez %>% 
    left_join(suitable_area, by= "rgn") %>% 
    mutate(zone_pct = suitable_area_km2 / area_km2 * 100)
  
  # Map for aquaculture sites of species 
aqua_map <- basemap + tm_shape(eez_aqua) +  # eez zones 
  tm_polygons(col = "zone_pct",
              palette = "Oranges", 
              alpha = 0.5,
              title = "% of suitable area \nper EEZ (km2)") + 
  tm_shape(suitable_aqua) + # suitable area
  tm_raster(palette = "firebrick", legend.show = FALSE) +
  tm_add_legend(type = "fill", 
                title = "Total suitable area (km^2)", 
                col = "firebrick", 
                labels = " ") +
  tm_layout(legend.outside = TRUE,
            fontfamily = "serif",
            main.title = paste(species_name,"Aquaculture \nSuitable Sites"))

  # table 
  eez_table <- eez_aqua %>% # Filter data for table 
    st_drop_geometry() %>% 
    dplyr::select(rgn, area_km2, suitable_area_km2, zone_pct)
  
   eez_table_html <- kable(eez_table,
        col.names = c("EEZ Region",
                      "EEZ total area (km^2)",
                      "Suitable aquaculture area (km^2)",
                      "% of suitable area per zone"),
        caption = paste("Percentage of suitable ", species_name, " aquaculture sites per EEZ"),
        digits = 2) %>% 
    kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center", html_font='times new roman')

  print(aqua_map)  
  eez_table_html  
}

```


Now, I will use the function to plug in the depth and temperature specifications for sugar kelp 

- sea surface temperature: 10-15째C
- depth: 0-30 meters below sea level

```{r}
aqua_fun(depth_low = -30, depth_high = 0, temp_low = 10, temp_high = 15, species_name = "Sugar Kelp")
```


The map and table show that there are very little sites along the west coast that have the right conditions to grow sugar kelp. Sugar kelp needs very cold water to thrive. 

White Seabass  
- 8 - 24 degrees Celsius and occupy the water column from 
- 0 - 122 meters beneath the surface 

```{r}
aqua_fun(depth_low = -122, depth_high = 0, temp_low = 8, temp_high = 24, species_name = "White Seabass")
```


## Citations

1.  NOAA Coral Reef Watch. 2008-2012, updated daily. NOAA Coral Reef Watch Version 3.1 Daily 5km Satellite Regional Virtual Station Time Series Data for Southeast Florida, Mar. 12, 2013-Mar. 11, 2014. College Park, Maryland, USA: NOAA Coral Reef Watch. Data set accessed 2024-11-12 at https://coralreefwatch.noaa.gov/product/vs/data.php.

2.  GEBCO Compilation Group (2024) GEBCO 2024 Grid (doi:10.5285/1c44ce99-0a0d-5f4f-e063-7086abc0ea0f)

3.  Flanders Marine Institute (2024): MarineRegions.org. Available online at www.marineregions.org. Consulted on 2024-11-13.

4.  White, N. & Marshall, C.E. 2007. Saccharina latissima Sugar kelp. In Tyler-Walters H. Marine Life Information Network: Biology and Sensitivity Key Information Reviews, \[on-line\]. Plymouth: Marine Biological Association of the United Kingdom. \[cited 12-11-2024\]. Available from: https://www.marlin.ac.uk/species/detail/1375
